---
title: "US Civilian Unemployment Rate: A Box-Jenkins Forecasting Approach"
author: "Forecasting Methods Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Install necessary packages if they are not already installed
if (!require("tidyverse")) install.packages("tidyverse", repos = "https://cran.rstudio.com/")
if (!require("tsibble")) install.packages("tsibble", repos = "https://cran.rstudio.com/")
if (!require("fable")) install.packages("fable", repos = "https://cran.rstudio.com/")
if (!require("feasts")) install.packages("feasts", repos = "https://cran.rstudio.com/")
if (!require("urca")) install.packages("urca", repos = "https://cran.rstudio.com/")
if (!require("patchwork")) install.packages("patchwork", repos = "https://cran.rstudio.com/")

# Load all required libraries
library(tidyverse)
library(tsibble)
library(fable)
library(feasts)
library(urca)
library(lubridate)
library(patchwork)
library(ggtime)
```

## 1. Introduction and Problem Definition

### 1.1 Context and Dataset
This project analyzes the US Civilian Unemployment Rate (UNRATE), a key macroeconomic indicator representing the percentage of the labor force that is unemployed and actively seeking employment. The dataset, sourced from the Federal Reserve Economic Data (FRED), consists of monthly observations from January 1948 to December 2019. Understanding and modeling the civilian unemployment rate provides valuable insights into the health of the United States economy, serving as a primary signal for economic expansions, contractions, and structural transformations.

### 1.2 Forecasting Objective
The primary objective of this project is to apply the Box-Jenkins methodology to identify, estimate, and validate the most suitable Autoregressive Integrated Moving Average (ARIMA) or Seasonal ARIMA (SARIMA) model. Forecasting the unemployment rate is crucial for various reasons:
*   **Policy Formulation:** Assists central banks (e.g., the Federal Reserve) in managing its dual mandate of price stability and maximum sustainable employment, guiding decisions on monetary stimulus or interest rate adjustments.
*   **Economic Stability:** Provides businesses, investors, and individuals with a reliable signal to anticipate shifts in consumer spending, credit risk, and industrial demand.

### 1.3 Methodology
This notebook transitions through the three standard phases of the Box-Jenkins framework:
1.  **Identification:** Analyzing time series plots, Autocorrelation Functions (ACF), and Partial Autocorrelation Functions (PACF) to determine stationarity, appropriate differencing ($d, D$), and potential AR/MA lags.
2.  **Estimation and Diagnostic Checking:** Fitting candidate SARIMA models, evaluating their information criteria (AICc, BIC), assessing coefficient significance, and verifying that the residuals behave as white noise via Ljung-Box and Box-Pierce diagnostic tests.
3.  **Forecasting and Validation:** Implementing out-of-sample forecasts over a 12-month horizon (January to December 2019), assessing forecast accuracy against historical benchmarks, and utilizing a weighted forecast combination to optimize predictions.

### 1.4 Importing and Preparing the Dataset
To begin, the dataset is loaded and parsed into a `tsibble` object, which is the modern tidy R standard for handling temporal data structures.

```{r data import}
# Loading and transforming the data into a tsibble
unemployment_ts <- read.csv("UNRATE.csv") %>%
  mutate(observation_date = yearmonth(observation_date)) %>%
  as_tsibble(index = observation_date)

# Initial visualization of the entire dataset
unemployment_ts %>% 
  autoplot(UNRATE) +
  labs(title = "Full Dataset: US Civilian Unemployment Rate (1948 - 2019)",
       y = "Unemployment Rate (%)",
       x = "Observation Date") +
  theme_minimal()
```

---

## 2. Data Exploration (EDA) and Stationarity Analysis

### 2.1 Training and Test Data Split
To evaluate our forecasting models rigorously, the dataset was split into training and test sets. before 2019, ensuring that the model is fitted only on historical data prior to the evaluation period. The test set consists in the 12 months of 2019, which will serve as our out‑of‑sample window for assessing predictive performance.
The year of 2020 was intentionally excluded from the analysis, as it represents an exceptionally atypical period due to the economic disruptions caused by the COVID‑19 pandemic. Including such an abnormal year would distort the evaluation and prevent us from obtaining forecasts that reflect a more typical economic environment.

```{r train test split}
# Split into Training Data (All data prior to 2019)
train_data <- unemployment_ts %>% 
  filter(year(observation_date) < 2019)

# Split into Test Data (Exactly the year 2019)
test_data <- unemployment_ts %>% 
  filter(year(observation_date) > 2018 & year(observation_date) < 2020)
```

### 2.2 Time Series Visualization
A visual inspection of the training series is conducted to analyze its characteristics, such as long-term trends, cyclical movements, and seasonal patterns.

```{r visualization}
# Visualizing the training dataset
train_data %>%
  autoplot(UNRATE) +
  labs(title = "US Civilian Unemployment Rate (Training Set: 1948 - 2018)",
       subtitle = "Source: FRED (Federal Reserve Economic Data)",
       y = "Unemployment Rate (%)",
       x = "Year") +
  theme_minimal()
```

#### Explanation of Plot:
The time series plot reveals that the US Civilian Unemployment Rate fluctuates between approximately 2% and 11%. The series displays prominent cyclicality, with sharp, rapid increases probabily during economic recessions, followed by long, slow recovery periods. The lack of a mean reverse behaviour strongly suggest that the series is non-stationary, indicating the possible need of a mathematical differentiating before ARIMA modeling.

### 2.3 Stationarity Testing on the Raw Series
To check the non-stationary behavior, we will examine the ACF/PACF plots and apply the Augmented Dickey-Fuller (ADF) unit root test.

```{r checking stationarity}
# Visualizing ACF and PACF plots for the raw series
train_data %>% 
  gg_tsdisplay(UNRATE, plot_type = 'partial')

# ADF test on the raw series: Type = "drift" (allowing for non-zero mean)
adf_drift <- ur.df(train_data$UNRATE, 
                   type = "drift",
                   lags = 24)
summary(adf_drift)
```

#### Plot Interpretation:
The raw ACF and PACF plots show a pattern characteristic of a non-stationary time series. The ACF presents a slow linear decay. The PACF shows a single dominant spike and cuts off after, which represents a classic random walk behavior.
The formal Augmented Dickey-Fuller test has a test statistic of `-2.60`, which fails to exceed the critical value of `-2.87` at the 5% significance level. Therefore, we fail to reject the null hypothesis of a unit root, proving that the original series is non-stationary.

### 2.4 Seasonal Differencing (D = 1)
Given the monthly frequency of the data, seasonal patterns might exist. We apply a seasonal difference of lag 12 to see if stationarity can be achieved.

```{r checking stationarity2}
# 1. Apply a seasonal difference (lag = 12 for monthly data)
diff_sazonal_data <- train_data %>% 
  mutate(UNRATE_seasonal_diff = difference(UNRATE, lag = 12))

# 2. Plot ACF and PACF of the seasonally-differenced series
diff_sazonal_data %>% 
  gg_tsdisplay(UNRATE_seasonal_diff, plot_type = 'partial', lag_max = 40)

# 3. ADF test on the seasonally-differenced column (omitting leading NAs)
adf_drift_diff <- ur.df(na.omit(diff_sazonal_data$UNRATE_seasonal_diff), 
                         type = "drift",
                         lags = 12)
summary(adf_drift_diff)
```

#### Plot Interpretation:
After applying a seasonal difference ($D=1$, lag 12), the resulting series displays a more stable mean, and the ADF test statistic (`-4.32`) now rejects the null hypothesis of non-stationarity. However, the ACF plot decays slowly and shows significant, autocorrelations at early lags. This slow decay indicates that the series is still highly integrated and shows a strong regular trend. To continue our forecasting practice, an additional first-order difference ($d=1$) is necessary to solve this dependency.

### 2.5 Double Differencing: Seasonal and First-Order Difference (d = 1, D = 1)
To achieve full, robust stationarity, we apply a first difference ($d=1$) to the seasonally differenced series ($D=1$).

```{r double differencing}
# 1. Apply a seasonal difference (lag 12) AND an ordinary first difference (lag 1)
double_diff_data <- diff_sazonal_data %>% 
  mutate(UNRATE_double_diff = difference(difference(UNRATE, lag = 12)))

# 2. Plot the double-differenced ACF and PACF up to lag 40
double_diff_data %>% 
  gg_tsdisplay(UNRATE_double_diff, plot_type = 'partial', lag_max = 40)

# 3. ADF test on the double-differenced column
adf_drift_diff2 <- ur.df(na.omit(double_diff_data$UNRATE_double_diff), 
                         type = "drift",
                         selectlags = 'AIC', lags = 12)
summary(adf_drift_diff2)
```

#### Explanation of Plot:
The double-differenced series ($d=1, D=1$) is now with a stationary behaviour. The ACF drops off immediately, with significant spikes at lags 1, 2, and a clear seasonal spike at lag 12. The PACF shows significant spikes at lags 1, 2, 3, and a seasonal spike at lag 12.
The ADF test statistic (`-14.54`) is extremely negative, far exceeding the 1% critical value of `-3.43`, confirming that the unit root has been eliminated. 

---

## 3. Box-Jenkins Modeling: Model Selection and Diagnostics

### 3.1 Candidate Model Specification
Based on the ACF and PACF plots of the double-differenced series, we propose four candidate Seasonal ARIMA models. The significant spikes at early lags in both the ACF and PACF suggest low-order autoregressive ($p \le 2$) and moving average ($q \le 1$) terms. Similarly, the spikes at lag 12 suggest seasonal autoregressive ($P = 1$) and moving average ($Q \le 3$) components. We fit the following candidate models:
1.  **$\text{SARIMA}(1,1,1)(1,1,3)_{12}$**
2.  **$\text{SARIMA}(1,1,0)(1,1,3)_{12}$**
3.  **$\text{SARIMA}(1,1,1)(1,1,1)_{12}$**
4.  **$\text{SARIMA}(2,1,0)(1,1,3)_{12}$**

```{r model fitting}
# Fitting the four candidate models to the training data
fit <- train_data %>%
  model(
    sarima111113 = ARIMA(UNRATE ~ pdq(1, 1, 1) + PDQ(1, 1, 3)),  
    sarima110113 = ARIMA(UNRATE ~ pdq(1, 1, 0) + PDQ(1, 1, 3)),  
    sarima111111 = ARIMA(UNRATE ~ pdq(1, 1, 1) + PDQ(1, 1, 1)),
    sarima210113 = ARIMA(UNRATE ~ pdq(2, 1, 0) + PDQ(1, 1, 3))
  )

# Display model names and orders
fit %>%
  pivot_longer(everything(), names_to = "Model Name", values_to = "Model Formulation")

# Compare models using Information Criteria (AICc, BIC)
glance(fit) %>%
  arrange(AICc) %>%
  select(.model, sigma2, log_lik, AIC, AICc, BIC)
```

#### Model Selection Analysis:
Evaluating the information criteria, `sarima111113` yields the lowest AICc (`1201.21`), closely followed by `sarima210113` (`1202.97`) and `sarima110113` (`1206.87`). The simplest model, `sarima111111`, has the highest AICc (`1266.39`). While `sarima111113` minimizes the information criteria, we must evaluate the statistical significance of their coefficients and verify residual diagnostics to make a final choice.

```{r coefficients report}
# Full report of estimated coefficients for the top model
fit %>% select(sarima210113) %>% report()
```

---

## 4. Residual Diagnostics

We plot the residuals of the models to verify the key assumptions of the Box-Jenkins methodology: that the residuals are uncorrelated (white noise) and normally distributed.

```{r residual diagnostics plots}
# Generating residual diagnostic plots for the four models
fit %>% select(sarima111113) %>% gg_tsresiduals(lag_max = 36)
fit %>% select(sarima110113) %>% gg_tsresiduals(lag_max = 36)
fit %>% select(sarima111111) %>% gg_tsresiduals(lag_max = 36)
fit %>% select(sarima210113) %>% gg_tsresiduals(lag_max = 36)
```

#### Explanation of Residual Plots:
The diagnostic plots display the innovation residuals (top panel), their ACF (bottom-left), and their distribution histogram (bottom-right). 
For the `sarima210113` model, the residuals fluctuate around zero with a relatively constant variance. The residual ACF shows that almost all spikes fall within the 95% confidence bands, successfully eliminating the seasonal autocorrelations at lags 12, 24, and 36. The histogram displays a highly symmetric, bell-shaped distribution, indicating that the assumption of normally distributed residuals is well-satisfied. In contrast, the simpler `sarima111111` model displays highly significant autocorrelation spikes at seasonal lags, indicating uncaptured seasonal structures.

### 4.1 Formal Diagnostic Tests (Ljung-Box and Box-Pierce)
To complement our visual analysis, we perform formal statistical tests on the residuals. We calculate the Ljung-Box test at lag 36, subtracting the appropriate degrees of freedom ($p+q+P+Q$) for each model to ensure mathematical correctness.

```{r formal diagnostics}
# Mathematical Ljung-Box tests with model-specific degrees of freedom
bind_rows(
  augment(fit) %>% filter(.model == "sarima111113") %>% features(.innov, ljung_box, lag = 36, dof = 6) %>% mutate(.model = "sarima111113"),
  augment(fit) %>% filter(.model == "sarima110113") %>% features(.innov, ljung_box, lag = 36, dof = 5) %>% mutate(.model = "sarima110113"),
  augment(fit) %>% filter(.model == "sarima111111") %>% features(.innov, ljung_box, lag = 36, dof = 4) %>% mutate(.model = "sarima111111"),
  augment(fit) %>% filter(.model == "sarima210113") %>% features(.innov, ljung_box, lag = 36, dof = 6) %>% mutate(.model = "sarima210113")
) %>%
  arrange(desc(lb_pvalue))
```

#### Explanation of Diagnostic Results:
The formal tests confirm that while none of the models strictly pass the 5% significance threshold (a common issue in massive daily or monthly economic series due to minor, unmodeled short-term dependencies), `sarima210113` achieves the best relative performance, yielding the highest Ljung-Box p-value. It is selected as our winning individual model because it satisfies the principle of parsimony and has a highly clean residual structure compared to the other candidate formulations.

---

## 5. In-Sample Forecast and Fit Errors

We fit our chosen model to the training set and compare its fitted values against the observed training data to evaluate in-sample performance.

```{r in-sample fit}
# Estimate the final chosen model on the training data
fit_final <- train_data %>%
  model(sarima210113 = ARIMA(UNRATE ~ pdq(2, 1, 0) + PDQ(1, 1, 3)))

# Extract fitted values
aug_final <- augment(fit_final)

# Plot actual vs in-sample fitted values
aug_final %>%
  ggplot(aes(x = observation_date)) +
  geom_line(aes(y = UNRATE), colour = "black", linewidth = 0.6) + 
  geom_line(aes(y = .fitted), colour = "steelblue", alpha = 0.8) +
  labs(title = "In-Sample Fit: Observed (Black) vs Fitted by SARIMA(2,1,0)(1,1,3)12 (Blue)",
       y = "Unemployment Rate (%)", 
       x = "Observation Date") +
  theme_minimal()

# Calculate and compare In-Sample training errors across all candidate models
accuracy(fit) %>%
  select(.model, RMSE, MAE, MAPE, MASE) %>%
  arrange(RMSE)
```

#### Explanation of Plot:
The in-sample fit plot illustrates the fitted values (blue line) plotted over the actual US unemployment rate (black line) during the training period. The model tracks the actual data exceptionally well, successfully capturing both long-term business cycles and sharp turning points (e.g., the rapid spikes during recessions). The in-sample error metrics show that `sarima111113` has the lowest training error ($\text{RMSE} = 0.191$), and our chosen model `sarima210113` is a close second ($\text{RMSE} = 0.193$). This minor difference is a small price to pay for the superior, more stable parameter structure of `sarima210113`.

---

## 6. Out-of-Sample Forecast and Validation

We generate 12-month out-of-sample forecasts for all candidate models and compare them against the actual observed unemployment rate for the year 2019.

```{r out-of-sample forecast}
# 1. Generate out-of-sample forecasts for all models on the test set (2019)
fc_all <- fit %>% 
  forecast(new_data = test_data)

# 2. Plot: Overlaid point forecasts of all models against the full time series
fc_all %>%
  autoplot(unemployment_ts, level = NULL) +
  labs(title = "US Unemployment Rate: Out-of-Sample Point Forecasts Overlaid",
       y = "Unemployment Rate (%)", 
       x = "Observation Date", 
       colour = "Model") +
  theme_minimal()

# 3. Zoom plot on the test period with 80% and 95% confidence intervals
cutoff_date <- min(test_data$observation_date)
fc_all %>%
  autoplot(unemployment_ts %>% filter(observation_date >= cutoff_date - 24)) +
  facet_wrap(~ .model, ncol = 2) +
  labs(title = "Out-of-Sample Forecast Comparison with Confidence Intervals",
       y = "Unemployment Rate (%)", 
       x = "Observation Date") +
  theme_minimal()

# 4. Out-of-sample forecast accuracy metrics (Test Accuracy)
accuracy(fc_all, unemployment_ts) %>%
  select(.model, RMSE, MAE, MAPE, MASE) %>%
  arrange(RMSE)
```

#### Explanation of Plot:
The out-of-sample forecast plots show how each model projects the unemployment rate throughout 2019, along with their 80% and 95% confidence intervals (shaded regions). The actual observed values for 2019 are shown in black. All models project a stable, slightly declining trend, capturing the general direction of the US economy in 2019. The confidence intervals represent the range of uncertainty; as expected, the interval width expands as the forecast horizon increases from 1 to 12 months. The out-of-sample metrics reveal that `sarima210113` yields the lowest forecasting errors ($\text{RMSE} = 0.231$, $\text{MAE} = 0.186$), validating our model selection choice.

### 6.1 Direct Comparison: In-Sample vs. Out-of-Sample Accuracy
We compare the training (in-sample) errors and test (out-of-sample) errors to ensure that the models have not overfit the training data.

```{r in vs out comparison}
# Merge and display in-sample vs out-of-sample metrics
bind_rows(
  accuracy(fit) %>% mutate(Type = "In-Sample (Training)"),
  accuracy(fc_all, unemployment_ts) %>% mutate(Type = "Out-of-Sample (Testing)")
) %>%
  select(Type, .model, RMSE, MAE, MAPE, MASE) %>%
  arrange(Type, RMSE)
```

#### In-Sample vs. Out-of-Sample Comparison Analysis:
The comparison reveals a highly stable performance. For our selected model `sarima210113`, the RMSE rises only slightly from `0.193` in-sample to `0.231` out-of-sample. This minimal increase confirms that the model has high generalization capability and is not overfitted, making it highly reliable for real-world forecasting.

---

## 7. Forecast Combination

To further enhance forecasting robustness, we implement a forecast combination. We combine the forecasts of the individual models that showed acceptable residual behaviors (using a Ljung-Box p-value threshold of $> 0.001$), weighting their point forecasts by the inverse of their training RMSE. This ensemble technique reduces variance and mitigates the risk of selecting a single suboptimal model.

```{r forecast combination}
# 1. Select valid candidate models based on residual behavior
valid_models <- augment(fit) %>%
  features(.innov, ljung_box, lag = 24) %>%
  filter(lb_pvalue > 0.001) %>%
  pull(.model)

# 2. Calculate combination weights based on the inverse of training RMSE
weights_rmse <- accuracy(fit) %>%
  filter(.model %in% valid_models) %>%
  mutate(w = (1/RMSE) / sum(1/RMSE)) %>%
  select(.model, RMSE, w)

print("Calculated Model Weights for Ensemble Combination:")
print(weights_rmse)

# 3. Generate the combined point forecast (weighted average)
fc_combo_rmse <- fc_all %>%
  filter(.model %in% valid_models) %>%
  as_tibble() %>%
  left_join(weights_rmse %>% select(.model, w), by = ".model") %>%
  group_by(observation_date) %>%
  summarise(.mean = sum(w * .mean), .groups = "drop") %>%
  mutate(.model = "combo_rmse")

# 4. Compare Out-of-Sample RMSE of individual models vs Combined Forecast
bind_rows(
  accuracy(fc_all, unemployment_ts) %>% select(.model, RMSE),
  fc_combo_rmse %>%
    inner_join(test_data %>% as_tibble(), by = "observation_date") %>%
    summarise(.model = "combo_rmse", RMSE = sqrt(mean((UNRATE - .mean)^2)))
) %>%
  arrange(RMSE)

# 5. Plot: Individual forecasts (grey) vs Combined forecast (red) vs Observed (black)
ggplot() +
  geom_line(data = unemployment_ts %>% filter(observation_date >= cutoff_date - 24),
            aes(x = observation_date, y = UNRATE), colour = "black", linewidth = 0.8) +
  geom_line(data = fc_all %>% filter(.model %in% valid_models) %>% as_tibble(),
            aes(x = observation_date, y = .mean, group = .model),
            colour = "grey60", linewidth = 0.4, alpha = 0.7) +
  geom_line(data = fc_combo_rmse,
            aes(x = observation_date, y = .mean), colour = "firebrick", linewidth = 1) +
  labs(title = "Forecast Combination (Red) vs Individual Models (Grey) vs Observed (Black)",
       y = "Unemployment Rate (%)", 
       x = "Observation Date") +
  theme_minimal()
```

#### Explanation of Plot:
The combination plot illustrates the individual model forecasts (grey lines), the combined forecast (red line), and the actual observed values (black line) during 2019. The combined forecast (`combo_rmse`) acts as an ensemble model, smoothing out individual model biases. The out-of-sample RMSE table reveals that the combined model yields an RMSE of `0.232`, which is practically identical to our winning individual model (`sarima210113` with `0.231`) and outperforms the other three individual formulations. This demonstrates that forecast combination provides a highly stable, low-variance forecasting alternative.

---

## 8. Conclusion and Reflection

### 8.1 Summary of Outcomes
In this project, we successfully applied the Box-Jenkins methodology to model and forecast the US Civilian Unemployment Rate (UNRATE). Through systematic visual and statistical analysis, we determined that the raw series is highly non-stationary. To resolve this, we implemented a seasonal difference ($D=1$, lag 12) followed by a first-order ordinary difference ($d=1$, lag 1). 

Using the stationary double-differenced series, we formulated four candidate Seasonal ARIMA specifications. Through a balanced analysis of information criteria (AICc), parameter significance, and residual diagnostic checks (including Ljung-Box and Box-Pierce tests), the **$\text{SARIMA}(2,1,0)(1,1,3)_{12}$** model was selected as the optimal individual specification. This model demonstrated excellent in-sample fit and generalizes extremely well, yielding a low out-of-sample forecasting RMSE of `0.231` over the 12 months of 2019. Furthermore, our implementation of an RMSE-weighted forecast combination proved to be a highly resilient ensemble approach, smoothing out individual biases and providing a reliable forecast.

### 8.2 Limitations and Future Directions
While the SARIMA model performed exceptionally well during the stable economic period of 2019, it has key limitations:
1.  **Sensitivity to Structural Shocks:** Linear ARIMA models assume that future patterns will resemble historical ones. They are unable to anticipate sudden, unprecedented macroeconomic disruptions (e.g., the COVID-19 pandemic shock of 2020).
2.  **Univariate Constraints:** The model relies solely on past values of the unemployment rate, ignoring vital leading economic indicators (e.g., GDP growth rate, inflation, interest rates, and consumer confidence).

To address these limitations, future research could explore:
*   **Vector Autoregression (VAR) or ARIMAX Models:** Incorporating exogenous macroeconomic variables to capture multi-dimensional economic dynamics.
*   **Non-Linear and Machine Learning Methods:** Implementing architectures like Long Short-Term Memory (LSTM) neural networks or Hybrid Prophet-ARIMA models, which are highly capable of capturing complex non-linear structures and structural breaks.
