---
title: "Long-term Unemployment Rate Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!require("tidyverse")) install.packages("tidyverse", repos = "https://cran.rstudio.com/")
if (!require("tsibble")) install.packages("tsibble", repos = "https://cran.rstudio.com/")
if (!require("fable")) install.packages("fable", repos = "https://cran.rstudio.com/")
if (!require("feasts")) install.packages("feasts", repos = "https://cran.rstudio.com/")
if (!require("urca")) install.packages("urca", repos = "https://cran.rstudio.com/")
if (!require("patchwork")) install.packages("patchwork", repos = "https://cran.rstudio.com/")

library(tidyverse)
library(tsibble)
library(fable)
library(feasts)
library(urca)
library(lubridate)
library(patchwork)
library(ggtime)
```
## 1. Introduction and Problem Definition

### 1.1 Context and Dataset
This project analyzes the Civilian Unemployment Rate (UNRATE), a critical macroeconomic indicator representing the percentage of the labor force that is unemployed and actively seeking employment. The dataset, sourced from the Federal Reserve Economic Data (FRED), consists of monthly observations that provide insight into the health of the economy.

### 1.2 Forecasting Objective
The primary goal is to apply the Box-Jenkins methodology to identify the most suitable ARIMA or SARIMA model for this series. Forecasting unemployment is of paramount importance for: 
Policy Formulation: Assisting central banks and government bodies in making informed decisions regarding interest rates and fiscal stimulus. 
Economic Stability: Providing businesses and individuals with a metric to anticipate shifts in consumer spending and market demand.

### 1.3 Methodology
This notebook will transition from Identification (using ACF/PACF plots) to Estimation and Diagnostic Checking to ensure the residuals behave as white noise. The final model will be used to generate a 12-month forecast, the accuracy of which will be assessed against historical trends.

### 1.4 Importing and preparing the Dataset

```{r data import}
# Loading and transforming the data into a tsibble
unemployment_ts <- read.csv("UNRATE.csv") %>%
  mutate(observation_date = as.Date(observation_date)) %>%
  as_tsibble(index = observation_date)
# 2. Fill the "holes" (Implicit Gaps)

unemployment_ts_clean<-fill_gaps(unemployment_ts) %>%                             

# 3. Fill the empty values (Explicit NAs)
# We use 'down' to fill a missing month with the previous month's rate
  fill(UNRATE, .direction = "down")
```

## 2. Data Exploration (EDA)

### 2.2 Checking and removing Missing Values
```{r vchecking and filling missing values}
# Visualizing the Unemployment Rate series
sum(is.na(unemployment_ts_clean$UNRATE))

# Remove NAs to prevent errors in ur.df
data <- unemployment_ts_clean %>%
  filter(!is.na(UNRATE))
```

### 2.2 Time Series Visualization
The visualization of the US Civilian Unemployment Rate reveals a series with significant cyclicality and sudden shocks, particularly during global economic crises. The absence of a constant mean and the presence of long-term "drifts" suggest that the series is non-stationary
```{r visualization}
# Visualizing the Unemployment Rate series
data %>%
  autoplot(UNRATE) +
  labs(title = "US Civilian Unemployment Rate (Monthly)",
       subtitle = "Source: FRED",
       y = "Unemployment Rate (%)",
       x = "Year") +
  theme_minimal()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2.3 Formal Stationarity Verification
To determine the integration order ($d$) of the series, we employ the Augmented Dickey-Fuller (ADF) test following a general-to-specific strategy. Since the initial time plot does not display a clear deterministic trend pattern, the first approach will be to test a model including only a constant (drift).

Step 1: Drift Model

We begin with a specification that includes a constant term, which is appropriate for an unemployment rate that does not fluctuate around zero but lacks a permanent linear trend.

$H_0$: The series has a unit root (is non-stationary).

$H_1$: The series is trend-stationary.

Lag Selection Strategy:

Following the manual requirement, I start the test with 16 lags (accounting for the monthly nature of the data) and decrease the count one by one until the last lagged difference is statistically significant. I will identify significance by checking if the p-value is less than 0.05 (or the presence of stars * in the coefficient table).

```{r checking stationarity}
# ADF test: Type = "drift"
adf_drift <- ur.df(data$UNRATE, 
                          type = "drift", 
                          lags = 0)
summary(adf_drift)
```

Through this iterative process, we arrived at Lag 0 as the only significant specification for the auxiliary regression. This simplifies the procedure from an Augmented Dickey-Fuller to a standard Dickey-Fuller (DF) test, represented by the following auxiliary expression:$$\Delta y_t = \alpha + \gamma y_{t-1} + \epsilon_t$$

Final Decision on Stationarity with the lag length fixed at 0, the test results are as follows:

Test Statistic ($\tau_2$): $-3.81$

Critical Value (5%): Approximately $-2.87$

Since the test statistic is significantly smaller than the critical value ($-3.81 < -2.87$), we reject the null hypothesis ($H_0$). This provides sufficient evidence to conclude that the US Unemployment Rate series is stationary ($d=0$) when accounting for a drift term. Consequently, no differencing is required for the subsequent identification of ARIMA components.  

```{r checking stationarity}
# ADF test: Type = "drift"
# ACF and PACF plots for the level series (d=0)
data %>%
  gg_tsdisplay(UNRATE, plot_type = 'partial') +
  labs(title = "ACF and PACF: US Civilian Unemployment Rate",
       subtitle = "Identification of p and q parameters for a stationary series",
       y = "Value",
       x = "Lag")
```
