---
title: "Long-term Unemployment Rate Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!require("tidyverse")) install.packages("tidyverse", repos = "https://cran.rstudio.com/")
if (!require("tsibble")) install.packages("tsibble", repos = "https://cran.rstudio.com/")
if (!require("fable")) install.packages("fable", repos = "https://cran.rstudio.com/")
if (!require("feasts")) install.packages("feasts", repos = "https://cran.rstudio.com/")
if (!require("urca")) install.packages("urca", repos = "https://cran.rstudio.com/")
if (!require("patchwork")) install.packages("patchwork", repos = "https://cran.rstudio.com/")

library(tidyverse)
library(tsibble)
library(fable)
library(feasts)
library(urca)
library(lubridate)
library(patchwork)
library(ggtime)
```
## 1. Introduction and Problem Definition

### 1.1 Context and Dataset
This project analyzes the Civilian Unemployment Rate (UNRATE), a critical macroeconomic indicator representing the percentage of the labor force that is unemployed and actively seeking employment. The dataset, sourced from the Federal Reserve Economic Data (FRED), consists of monthly observations that provide insight into the health of the economy.

### 1.2 Forecasting Objective
The primary goal is to apply the Box-Jenkins methodology to identify the most suitable ARIMA or SARIMA model for this series. Forecasting unemployment is of paramount importance for: 
Policy Formulation: Assisting central banks and government bodies in making informed decisions regarding interest rates and fiscal stimulus. 
Economic Stability: Providing businesses and individuals with a metric to anticipate shifts in consumer spending and market demand.

### 1.3 Methodology
This notebook will transition from Identification (using ACF/PACF plots) to Estimation and Diagnostic Checking to ensure the residuals behave as white noise. The final model will be used to generate a 12-month forecast, the accuracy of which will be assessed against historical trends.

### 1.4 Importing and preparing the Dataset

```{r data import}
# Loading and transforming the data into a tsibble
unemployment_ts <- read.csv("UNRATE.csv") %>%
  mutate(observation_date = yearmonth(observation_date)) %>%
  as_tsibble(index = observation_date)

# 1. Clean the entire dataset first (Fill implicit holes and explicit NAs)
unemployment_ts %>% 
  autoplot(UNRATE)

# 2. Split into Training Data (All data prior to January 2019)

```

## 2. Data Exploration (EDA)

### 2.2 Checking and removing Missing Values

```



```{r}
train_data <- unemployment_ts %>% 
  filter(year(observation_date) < 2018)

# 3. Split into Test Data (Exactly the year 2019)
# Method A: Using exact date boundaries
test_data <- data %>% 
  filter(year(observation_date) > 2018 & year(observation_date) < 2020)

# Alternative Method B (Cleaner): Using lubridate's year() function
# test_data <- unemployment_clean %>% filter(year(observation_date) == 2019)
```

### 2.2 Time Series Visualization
The visualization of the US Civilian Unemployment Rate reveals a series with significant cyclicality and sudden shocks, particularly during global economic crises. The absence of a constant mean and the presence of long-term "drifts" suggest that the series is non-stationary
```{r visualization}
# Visualizing the Unemployment Rate series
train_data %>%
  autoplot(UNRATE) +
  labs(title = "US Civilian Unemployment Rate (Monthly)",
       subtitle = "Source: FRED",
       y = "Unemployment Rate (%)",
       x = "Year") +
  theme_minimal()
```

GTS approach to find a significant lag

```{r checking stationarity}
train_data %>% 
  gg_tsdisplay(UNRATE, plot_type='partial')


# ADF test: Type = "drift"
adf_drift <- ur.df(train_data $UNRATE, 
                          type = "drift",
                          lags=24)
summary(adf_drift)

```

A seguir vamos diferenciar 

```{r checking stationarity}
# 1. Criar o dataset com a variável diferenciada
# Isto sim, é uma diferenciação sazonal (ex: lag = 12 para dados mensais)
diff_sazonal_data <- train_data %>% 
  mutate(UNRATE_seasonal_diff = difference(UNRATE, lag = 12))
# 2. Plotar a ACF e PACF da série diferenciada
diff_sazonal_data %>% 
  gg_tsdisplay(UNRATE_seasonal_diff, plot_type = 'partial', lag_max = 40)
```

Diferenciamos e agora vamos fazer de novo o teste D=1
```{r}
# ADF test: Type = "drift"
adf_drift_diff <- ur.df(diff_sazonal_data $UNRATE, 
                          type = "drift",
                          lags=12)
summary(adf_drift_diff)
```

We find significance in lag 12 and reject the null having the ts as stationary but the ac f and pacf show non-stationarity. 
This is due to the fact that we are using percentual data per se and bec ause it is not infitine it doesnt behave as a Random Walk na definição teórica mais clássica.
```{r}
# 1. Apply a seasonal difference (lag 12) AND an ordinary difference (lag 1)
double_diff_data <- diff_sazonal_data %>% 
  mutate(UNRATE_double_diff = difference(difference(UNRATE, lag = 12)))

# 2. Plot the new ACF and PACF up to lag 40
double_diff_data %>% 
  gg_tsdisplay(UNRATE_double_diff, plot_type = 'partial', lag_max = 40)
```
Now the ts looks finally stationary!!!!!! D=2
we will do the DF test again
```{r}
adf_drift_diff2 <- ur.df(double_diff_data $UNRATE, 
                          type = "drift",
                          lags=12)
summary(adf_drift_diff2)
```
Significance in lag 12 and the test result forces us to reject H0 so proves stationarity but because it gives us exacly the same statistical values as the cell of the 1st diff i decided to try another approac h and it gave the same result so now our TS is stationary

```{r}

adf_drift_diff2 <- ur.df(double_diff_data $UNRATE, 
                          type = "drift",
                          selectlags='AIC', lags=12)
summary(adf_drift_diff2)

```
time to classify the model:
SARIMA(1,0,1)(1,2,3)12
SARIMA(1,0,0)(1,2,3)12
